---
title: "tidy-model"
author: "Maximus Walton"
date: "2025-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyverse') ; library('modeldata')
```

## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.

```{r load data, include = FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?
In this report, we examine what factors predict a cricket's chirp rate.

We can see in the data that there is 31 observations, 2 species, the temperature ranges from 17.20 ti 39,49 and the mean chirping rate is 72.89.

```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```
## Temperature affects chirp rate

```{r temp, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```
Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
  
species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```
From the graph, we can see that as temperature increases, the chirping rate also increases for both species. However, there is a noticeable difference between the two species—O. exclamationis tends to chirp at a higher rate compared to O. niveus, even at the same temperature.

## Interactions

```{r species historgram, echo = FALSE}
ggplot(crickets, aes(x = rate, fill = species)) +
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```

```{r interactions, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets) 
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
```

##Visualize and model diamond prices

```{r loadDiamonds, include = FALSE}
data(diamonds, package = "ggplot2")

summary(diamonds)
```

```{r Diamonds, echo=FALSE}
set.seed(123)  # Ensures reproducibility
diamond_sample <- diamonds[sample(nrow(diamonds), 5000), ]  # Randomly sample 5000 points

ggplot(diamond_sample, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Effect of Carat & Cut on Diamond Price (Sampled Data)",
       x = "Carat",
       y = "Price")

diamond_lm <- lm(price ~ carat + cut + clarity, data = diamonds)
summary(diamond_lm)
```













